---
title: 'Class X: Topic'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{maynooth_uni_logo.jpg}
  \newline PRESS RECORD 
  \newline \vspace{1cm}
  \newline https://andrewcparnell.github.io/intermediate_ML
output:
  beamer_presentation:
    includes:
      in_header: header.tex
classoption: "aspectratio=169"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 5)
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
library(tidyverse)
```

## Things still to cover

-  Silhouette scores
-  'Add' layers?
-  This: https://tensorflow.rstudio.com/examples/
-  Saving and loading a keras object
-  Transfer learning
-  Model tuning
-  Things we don't cover 
	-  Running models on a GPU
	-  Basics of tensorflow
	-  Double descent
	-  Mathematics of these models
	-  GANs
-  Exercises
	-  Compare BART, BASS and Bayesian GPs
-  callbacks?
-  UMAP
-  Saving and loading keras models
-  Optimisation functions: what is Adam etc?

## Slide

```{r}
# Code
```

## Picture

\centering\includegraphics[width=3cm]{maynooth_uni_logo.jpg}

\let\thefootnote\relax\footnotetext{\tiny From URL}

